SU($d$) spin model dynamics

### 2020-10-28 sun_phases/sun_dynamics_notes.tex
minor changes to the 'simple model' of two large spins

Here we consider a relatively simple Hamiltonian on $2N$ spins,
\begin{align}
  H = \f{U}{2} \v S\c\v S + \sum_{j=1}^{2N} h_j s_{\z,j},
  &&
  h_j \equiv
  \begin{cases}
    +1 & j \le N \\
    -1 & j > N
  \end{cases}.
  \label{eq:two_spins_start}
\end{align}
Defining collective the spin-$\alpha$ operators for each sub-ensemble indexed by $k\in\set{0,1}$,
\begin{align}
  S_{\alpha k} = \sum_{j=1+kN}^{N+kN} s_{\alpha j},
\end{align}
as long as each sub-ensemble remains within the manifold of spin-polarized states we can write the Hamiltonian in \eqref{eq:two_spins_start} in the form
\begin{align}
  H = U \v S_0 \c\v S_1 + S_{\z,0} - S_{\z,1}
  = \sum_{m,n=-IN}^{IN} \p{U\op{mn}{nm}
    + \p{m-n}\op{mn}},
\end{align}
where $m,n$ are projections of a net sub-ensemble spin onto the $z$ axis, and we have neglected the scalars $\v S_k\cdot \v S_k = IN\p{IN+1}$.
Indexing states $\ket{mn}$ by the mean spin projection $\bar m=\p{m+n}/2$ and spin imbalance $\delta=m-n$,
\begin{align}
  \ket{\bar m\delta} \equiv \ket{m+\delta/2,m-\delta/2},
\end{align}
we can expand
\begin{align}
  \v S_0\c \v S_1 = \sum_{m,\delta}
  \op{\bar m\delta}{\bar m,-\delta},
  &&
  S_{\z,0} - S_{\z,1} = \sum_{m,\delta} \delta \op{\bar m\delta},
\end{align}
which allows us to write the Hamiltonian in the form
\begin{align}
  H = \sum_m \op{\bar m} \otimes
  \sum_\delta \p{U \op{\delta}{-\delta} + \delta \op{\delta}}.
\end{align}
Diagonalizing this Hamiltonian thus reduces to a qubit problem, for which we define the Pauli operator
\begin{align}
  X_\delta\p{U} \equiv \sum_{\sigma\in\set{\pm1}}
  \sigma \op{\delta,U,\sigma},
\end{align}
in terms of the rotated states
\begin{align}
  \ket{\delta,U,\sigma} \equiv \cos\p{\f{\eta}{2}} \ket{\p{\sigma\delta}}
  + \sigma \sin\p{\f{\eta}{2}} \ket{\p{-\sigma\delta}},
  &&
  \eta \equiv \arctan\p{\f{U}{\delta}},
\end{align}
defined for $\delta\ne0$.
Altogether, the Hamiltonian is then
\begin{align}
  H = \sum_m \op{\bar m} \otimes
  \sp{U \op{0} + \sum_{\delta>0} U_\delta X_\delta\p{U}}
  &&
  U_\delta \equiv \sqrt{U^2+\delta^2}.
\end{align}
The eigenvalues of $H$ are therefore $U$ and $\pm U_\delta$ for integer $\delta>0$.
We now consider an initial state in which both large spins point at a polar angle $\theta$ and azimuthal angles $\pm\phi/2$:
\begin{align}
  \ket{\theta,\phi} \equiv \sum_{m,n}
  \sqrt{\chi_m\p{\theta} \chi_n\p{\theta}} \,
  e^{-\i\p{m-n}\phi} \ket{mn}
  = \sum_{m,\delta}
  \chi_{m\delta}\p{\theta} e^{-\i\delta\phi}
  \ket{\bar m\delta}.
\end{align}
where
\begin{align}
  \chi_m\p\theta \equiv { 2IN \choose IN + m }
  \cos\p{\f{\theta}{2}}^{2IN+2m} \sin\p{\f{\theta}{2}}^{2IN-2m},
  &&
  \chi_{m\delta}\p\theta
  \equiv \sqrt{\chi_{m+\delta/2}\p\theta \chi_{m-\delta/2}\p\theta}.
\end{align}
In the eigenbasis of $H$, we have
\begin{align}
  \ket{\theta,\phi} = \sum_m \ket{\bar m} \otimes
  \sp{\chi_0\p\theta \ket{0}
    + \sum_{\delta>0} \chi_{m\delta}\p\theta
    \sum_{\sigma\in\set{\pm1}} e^{-\i\sigma\delta\phi}
    \sp{\cos\p{\f{\eta}{2}} \ket{\delta,U,\sigma}
    - \sigma\sin\p{\f{\eta}{2}} \ket{\delta,U,-\sigma}}},
\end{align}
where again $\eta\equiv\arctan\p{U/\delta}$ for shorthand.
Note that the only dependence of this state on the nuclear spin $I$ appears in the range of $\bar m$, so any distinction between observed behavior at different $I$ can essentially be attributed to the measurement of different observables.

We now turn our attention to the measurement of collective observables.
In particular, we consider the collective $N$-spin operators
\begin{align}
  S_{\mu\nu} \equiv \sum_{j=1}^N s_{\mu\nu j},
\end{align}
where $\mu,\nu\in\set{-I,-I+1,\cdots,I}$ and $s_{\mu\nu j} = \op{\mu}{\nu}_j$ is a spin-transition operator for spin $j$.
Rather than initial/final spin projections $\mu\to\nu$, it will be convenient to consider operators that induce spin transitions $\mu\to\mu+\epsilon$, and then expand these operators in the Dicke basis for spin-polarized states:
\begin{align}
  \tilde S_{\mu\epsilon} \equiv S_{\mu+\epsilon,\mu}
  = \sum_m \xi_{\mu\epsilon}^m \op{m+\epsilon}{m},
  &&
  \xi_{\mu\epsilon}^m \equiv \bk{m+\epsilon|S_{\mu+\epsilon,\mu}|m}.
\end{align}
In turn, we symmetrize these operators over their action on two sub-ensembles of $N$ spins each:
\begin{align}
  S_{\mu\epsilon\sigma} \equiv \tilde S_{\mu\epsilon} \otimes \1
  + \sigma \1 \otimes \tilde S_{\mu\epsilon}
  = \sum_{m,n} \xi_{\mu\epsilon}^m
  \p{\op{m+\epsilon,n}{mn} + \sigma \op{n,m+\epsilon}{nm}},
\end{align}
where $\sigma\in\set{\pm1}$.
In the symmetrized basis of a mean spin and spin imbalance, we have that
\begin{align}
  S_{\mu\epsilon\sigma}= \sum_{m,\delta,\sigma} \p{\xi_{\mu\epsilon}^{m+\delta/2}
    \ket{\bar m_\epsilon,\delta+\epsilon}
    + \sigma \xi_{\mu\epsilon}^{m-\delta/2}
    \ket{\bar m_\epsilon,\delta-\epsilon}}
  \bra{\bar m\delta},
  &&
  \bar m_\epsilon \equiv \bar m + \epsilon/2.
\end{align}

>>>>

\subsection{Hamiltonian}

Here we consider a simple Hamiltonian on $2N$ spins,
\begin{align}
  H = \f{U}{2} \v S\c\v S + \sum_{j=1}^{2N} h_j s_{\z,j},
  &&
  h_j \equiv
  \begin{cases}
    +1 & j \le N \\
    -1 & j > N
  \end{cases}.
  \label{eq:two_spins_start}
\end{align}
Defining collective the spin-$\alpha$ operators for each sub-ensemble indexed by $k\in\set{0,1}$,
\begin{align}
  S_{\alpha k} = \sum_{j=1+kN}^{N+kN} s_{\alpha j},
\end{align}
as long as each sub-ensemble remains within the manifold of spin-polarized states we can write the Hamiltonian in \eqref{eq:two_spins_start} in the form
\begin{align}
  H = U \v S_0 \c\v S_1 + S_{\z,0} - S_{\z,1}
  = \sum_{m,n=-IN}^{IN} \p{U\op{mn}{nm}
    + \p{m-n}\op{mn}},
\end{align}
where $m,n$ are projections of a net sub-ensemble spin onto the $z$ axis, and we have neglected the scalars $\v S_k\cdot \v S_k = IN\p{IN+1}$.
Indexing states $\ket{mn}$ by the mean spin projection $\bar m=\p{m+n}/2$ and spin imbalance $\delta=m-n$,
\begin{align}
  \ket{\bar m\delta} \equiv \ket{m+\delta/2,m-\delta/2},
\end{align}
we can expand
\begin{align}
  \v S_0\c \v S_1 = \sum_{m,\delta}
  \op{\bar m,-\delta}{\bar m\delta},
  &&
  S_{\z,0} - S_{\z,1} = \sum_{m,\delta} \delta \op{\bar m\delta},
\end{align}
which allows us to write the Hamiltonian in the form
\begin{align}
  H = \sum_m \op{\bar m} \otimes
  \sum_\delta \p{U \op{-\delta}{\delta} + \delta \op{\delta}}.
\end{align}
Diagonalizing this Hamiltonian thus reduces to a qubit problem, for which we define the Pauli operator
\begin{align}
  X_\delta\p{U} \equiv \sum_{\sigma\in\set{\pm1}}
  \sigma \op{\delta,U,\sigma},
\end{align}
in terms of the rotated states
\begin{align}
  \ket{\delta,U,\sigma} \equiv \cos\p{\f{\eta}{2}} \ket{\p{\sigma\delta}}
  + \sigma \sin\p{\f{\eta}{2}} \ket{\p{-\sigma\delta}},
  &&
  \eta \equiv \arctan\p{\f{U}{\delta}},
\end{align}
defined for $\delta\ne0$.
In words, the state $\ket{\delta,U,\pm}$ points in the direction of $\pm\p{\delta\uv{z}+U\uv{x}}$ on the Bloch sphere defined by the state $\set{\ket{\delta},\ket{-\delta}}$.
Altogether, the Hamiltonian is
\begin{align}
  H = \sum_m \op{\bar m} \otimes
  \sp{U\op{0} + \sum_{\delta>0} U_\delta X_\delta\p{U}},
  &&
  U_\delta \equiv \sqrt{U^2+\delta^2},
\end{align}
which has eigenvalues $U$ and $\pm U_\delta$ for integer $\delta>0$.

\subsection{Initial state}

We now consider an initial state with both large spins polarized at a polar angle $\theta$ and azimuthal angles $\pm\phi/2$:
\begin{align}
  \ket{\theta,\phi} \equiv \sum_{m,n}
  \sqrt{\chi_m\p{\theta} \chi_n\p{\theta}} \,
  e^{-\i\p{m-n}\phi} \ket{mn}
  = \sum_{m,\delta}
  \chi_{m\delta}\p{\theta} e^{-\i\delta\phi}
  \ket{\bar m\delta}.
\end{align}
where
\begin{align}
  \chi_m\p\theta \equiv { 2IN \choose IN + m }
  \cos\p{\f{\theta}{2}}^{2IN+2m} \sin\p{\f{\theta}{2}}^{2IN-2m},
  &&
  \chi_{m\delta}\p\theta
  \equiv \sqrt{\chi_{m+\delta/2}\p\theta \chi_{m-\delta/2}\p\theta}.
\end{align}
In the eigenbasis of $H$, we have
\begin{align}
  \ket{\theta,\phi} = \sum_m \ket{\bar m} \otimes
  \sp{\chi_0\p\theta \ket{0}
    + \sum_{\delta>0} \chi_{m\delta}\p\theta
    \sum_{\sigma\in\set{\pm1}} e^{-\i\sigma\delta\phi}
    \sp{\cos\p{\f{\eta}{2}} \ket{\delta,U,\sigma}
    - \sigma\sin\p{\f{\eta}{2}} \ket{\delta,U,-\sigma}}},
\end{align}
where again $\eta\equiv\arctan\p{U/\delta}$ for shorthand.
Note that the only dependence of this state on the nuclear spin $I$ appears in the range of $\bar m$, so any distinction between observed behavior at different $I$ can essentially be attributed to the measurement of different observables.

\subsection{Observables}

We now turn our attention to the measurement of collective observables.
In particular, we consider the collective $N$-spin operators
\begin{align}
  S_{\mu\nu} \equiv \sum_{j=1}^N s_{\mu\nu j},
\end{align}
where $\mu,\nu\in\set{-I,-I+1,\cdots,I}$ and $s_{\mu\nu j} = \op{\mu}{\nu}_j$ is a spin-transition operator for spin $j$.
Rather than initial/final spin projections $\nu\to\mu$, it will be convenient to consider operators that induce spin transitions $\mu\to\mu+\epsilon$, and then expand these operators in the spin-projection (Dicke) basis for spin-polarized states:
\begin{align}
  \tilde S_{\mu\epsilon} \equiv S_{\mu+\epsilon,\mu}
  = \sum_m \xi_{\mu\epsilon}^m \op{m+\epsilon}{m},
  &&
  \xi_{\mu\epsilon}^m \equiv \bk{m+\epsilon|S_{\mu+\epsilon,\mu}|m}.
\end{align}
In turn, we symmetrize these operators over their action on two sub-ensembles of $N$ spins each:
\begin{align}
  \tilde S_{\mu\epsilon}^\sigma \equiv \tilde S_{\mu\epsilon} \otimes \1
  + \sigma \1 \otimes \tilde S_{\mu\epsilon}
  = \sum_{m,n} \p{\xi_{\mu\epsilon}^m \ket{m+\epsilon,n}
    + \sigma \xi_{\mu\epsilon}^n \ket{m,n+\epsilon}} \bra{mn},
\end{align}
where $\sigma\in\set{\pm1}$.
In the symmetrized basis of a mean spin and spin imbalance, we have that
\begin{align}
  \tilde S_{\mu\epsilon}^\sigma
  = \sum_{m,\delta,\sigma} \p{\xi_{\mu\epsilon}^{m+\delta/2}
    \ket{\bar m_\epsilon,\delta+\epsilon}
    + \sigma \xi_{\mu\epsilon}^{m-\delta/2}
    \ket{\bar m_\epsilon,\delta-\epsilon}}
  \bra{\bar m\delta},
  &&
  \bar m_\epsilon \equiv \bar m + \epsilon/2.
\end{align}
